---
title: "R crash course goals"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Goals

Things to be able to do:

###1) read in your data

```{r}
batdat = read.csv("bat_data.csv")
#or 
#batdat = read.csv("Users/klangwig/Dropbox/teaching/bat_data.csv")

```

###2) load and install packages 
```{r}

#install.packages("ggplot2") #install a package (or use GUI system)

library(ggplot2) #load a package before use
```

###3) get help
```{r}
?"unique"
```


###4) examine your data

```{r, results='hide'}
unique(batdat$species)
str(batdat)
head(batdat)
tail(batdat)
dim(batdat)
names(batdat)
nrow(batdat)
ncol(batdat)
```

###5) do a calculation with your data 

```{r, results='hide'}
log(batdat$gdL)
log10(batdat$gdL)
3+3
batdat$temp + 10
```

###6) make a new column
```{r}
batdat$log.loads = log10(batdat$gdL) #make a new column that is the log of this column

```

###7) subset data - using subset(), or square brackets []
```{r}
#using subset
MYSE.dat = subset(batdat, species=="MYSE") #a factor/character, so need == and quotes
dim(MYSE.dat) #what are the dimensions of the new data frame?

warm.temps = subset(batdat, temp>6) #a number, so no quotes
dim(warm.temps)

#same thing using square brackets (say "where")
MYSE.dat = batdat[batdat$species=="MYSE",]
dim(MYSE.dat)

warm.temps = batdat[batdat$temp>6,]
```

###8) use aggregate or table to summarize some data
```{r}
bat.summary = aggregate(log.loads~species, FUN=mean, data = batdat) #aggregate data using the mean to give fungal loads by species
bat.summary
```

###9) make a histogram of a column in your data
```{r}
hist(batdat$log.loads)
```

###10) write out a dataframe
```{r}
write.csv(bat.summary, "bat.summary.csv",row.names = F)

```
  
###11) save script, close R **without saving workspace**
  
###12) re-open R, repeat the same thing by re-running your script after clearing your workspace


###Helpful links:  
https://greggilbertlab.sites.ucsc.edu/teaching/rtransition/

https://www.statmethods.net/

